### 2016-07-05 store-consistency.tex
C --> B. KC's suggestions incorporated.

\begin{figure}
\centering
\begin{txnimpcode}
$\begin{decoration}
  \hspace*{0.3in}P : \{\neg\committed(\C{Wd1}) \conj \neg\committed(\C{Wd2}) \conj
                          \C{C} = \C{k}\}
\end{decoration}$

 $\begin{decoration}
 P_1:\{ \neg\committed(\C{Wd1}) \conj
        (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k}) \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow 
            \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
              $\vdots$ 
     $\begin{decoration}
     \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
        \hspace*{0.3in}({\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \wrstoar \C{C}) \conj \neg\committed(\C{Wd1}) \conj \\
        \hspace*{0.32in}{\committed(\C{Wd2})} \wedge
         {\C{Wd2} \visar \C{Wd1}} 
        \Rightarrow \C{C = k-a2} \wedge \\
        \hspace*{1.9in}\C{v3 = k-a2-a1}\}
     \end{decoration}$ 
     C := v3
     $\begin{decoration}
      \phi_5 : \{({\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
          \conj \neg\committed(\C{Wd1}) \conj\\
         \hspace*{0.4in}{\committed(\C{Wd2})} 
            \Rightarrow \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a1-a2}\}
      \end{decoration}$ 
              $\vdots$ 
  }
 $\begin{decoration}
  Q_1 : \{{(\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1})
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.4in}{\committed(\C{Wd2})} 
          \Rightarrow \C{C = k-a1-a2} \}
  \end{decoration}$

$\begin{decoration}
  \hspace*{0.3in}Q : \{\committed(\C{Wd1}) \conj \committed(\C{Wd2}) \conj
                          \C{C} = \C{k-a1-a2}\}
\end{decoration}$
\end{txnimpcode}
\vspace*{-8pt}
\caption{\small \C{Wd1} transaction decorated with assertions}
\label{fig:wd1-decorated}
\vspace*{-12pt}
\end{figure}

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{The supplementary material shows the fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{C}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
$Q_1$ of \C{Wd1} asserts different values for \C{C} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. We consider the
triple associated with the assignment to \C{C} for illustration. The
pre-condition $\phi_4$ asserts different values for \C{C} and \C{v3}
(temporary alias of RHS expression) depending on whether or not
\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
this is sufficient to prove the post-condition $\phi_5$, which appears
to make a stronger claim about the value of \C{C}. The proof follows
from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
to easily conclude the following about the execution state after the
assignment:\vspace*{-12pt}

\begin{smathpar}
\begin{array}{l}
    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C} 
    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) \conj\\
    ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
    \C{C}) \conj \\
    ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
    \Rightarrow \C{C = k-a2} \wedge \C{C = k-a2-a1})
\end{array}
\end{smathpar}

\noindent The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions now write to \C{C}, {\sc si} requires either
$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
subsumes {\sc rc}) prohibits the former possibility, allowing us the
deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
${\committed(\C{Wd2})}  \Rightarrow \C{C = k-a2} \wedge \C{C =
k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.

The second part of the proof is to show that assertions are stable
despite the interference from the the concurrent thread executing
\C{Wd2}. We formalize the interference as a rely relation that
describes the semantics of \C{Wd2} at just the right level of
abstraction. In particular, we would like to capture two important
facts about \C{Wd2}: 
\begin{itemize}
  \item  If \C{Wd2} commits, then it should have already written to
  \C{C}. This fact is needed to show that \C{Wd2} conflicts with
  \C{Wd1}, so that \C{SI(Wd1)} property can be exploited to deduce
  $\C{Wd2} \visar \C{Wd1}$ from $\committed(\C{Wd2})$ in the
  sequential proof.

  \item The value written to \C{C} is either $\C{k-a1-a2}$ or
  $\C{k-a1}$ depending on whether or not \C{Wd1} has already
  committed. This is clearly necessary to deduce the value of \C{C}
  before and after \C{Wd1} commits.
\end{itemize}
 The resultant rely relation ($R_1$) is shown below (note: we
write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express that \C{Wd2}
is not committed in $\E$, but committed in $\E'$.):

\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \neg\underE{\committed(\C{Wd2})} \conj \I(\E) \conj \I(\E')\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
%   & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
%       (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \\
    & & \hspace*{0.37in} \conj \underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \C{COMMIT(Wd2)} \in (\E'.\A-\E.\A) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \\
    & & \hspace*{0.37in}\conj \forall(\eta \in \E'.\A - 
        \E.\A).~\txn(\eta) = \C{Wd2}\}\\
\end{array}
\end{smathpar}
For the sake of completeness, we also explicitly represent the fact
that all the effects added by $R_1$ are of \C{Wd2}. This is needed to
prove that an interference from \C{Wd2} does not affect the local
state of \C{Wd1}.

The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
$R_1$ does not commit \C{Wd2}, then the value of \C{C} is unaffected,
and if it commits \C{Wd2}, then $\C{Wd2}\not\visar\C{Wd1}$, hence
$\phi_4$ doesn't care. For $\phi_5$, {\sc si} condition should be used
to show that any interference from \C{Wd2} is invalid. The proof
proceeds on the same lines as the Hoare proof discussed above; it is
also discussed in detail in \S\ref{sec:motivation}.

>>>>

\begin{figure}
\centering
\begin{txnimpcode}
$\begin{decoration}
  \hspace*{0.3in}P : \{\neg\committed(\C{Wd1}) \conj \neg\committed(\C{Wd2}) \conj
                          \C{B} = \C{k}\}
\end{decoration}$

 $\begin{decoration}
 P_1:\{ \neg\committed(\C{Wd1}) \conj
        (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k}) \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow 
            \C{Wd2} \wrstoar \C{B} \wedge \C{B = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
              $\vdots$ 
     $\begin{decoration}
     \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{B = k} \wedge \C{v3 = k-a1} \conj\\
        \hspace*{0.3in}({\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \wrstoar \C{B}) \conj \neg\committed(\C{Wd1}) \conj \\
        \hspace*{0.32in}{\committed(\C{Wd2})} \wedge
         {\C{Wd2} \visar \C{Wd1}} 
        \Rightarrow \C{B = k-a2} \wedge \\
        \hspace*{1.9in}\C{v3 = k-a2-a1}\}
     \end{decoration}$ 
     B := v3
     $\begin{decoration}
      \phi_5 : \{({\neg\committed(\C{Wd2})} \Rightarrow \C{B = k-a1}) 
          \conj \neg\committed(\C{Wd1}) \conj\\
         \hspace*{0.4in}{\committed(\C{Wd2})} 
            \Rightarrow \C{Wd2} \wrstoar \C{B} \wedge \C{B = k-a1-a2}\}
      \end{decoration}$ 
              $\vdots$ 
  }
 $\begin{decoration}
  Q_1 : \{{(\neg\committed(\C{Wd2})} \Rightarrow \C{B = k-a1})
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.4in}{\committed(\C{Wd2})} 
          \Rightarrow \C{B = k-a1-a2} \}
  \end{decoration}$

$\begin{decoration}
  \hspace*{0.3in}Q : \{\committed(\C{Wd1}) \conj \committed(\C{Wd2}) \conj
                          \C{B} = \C{k-a1-a2}\}
\end{decoration}$
\end{txnimpcode}
\vspace*{-8pt}
\caption{\small \C{Wd1} transaction decorated with assertions}
\label{fig:wd1-decorated}
\vspace*{-12pt}
\end{figure}

Decorated parts of \C{Wd1} relevant to the discussion are shown in
Fig.~\ref{fig:wd1-decorated}.  \C{Wd2} is similar and not
shown\footnote{The supplementary material shows the fully decorated program.}.
Temporary name \C{v3} has been introduced to capture the result of the
RHS expression \C{C-a1}.  Assertions implicitly refer to the current
execution ($\E$), just as Hoare triples implicitly refer to the
current state.  The context for propositions is also the implicit,
i.e., we write $\psi$ instead of $\underE{\psi}$. The proposition \C{k
$\ge$ a1+a2} remains an invariant, hence elided. The pre-condition
$P_1$ of \C{Wd1} accounts for the possibility of \C{Wd2} committing
before \C{Wd1}, writing \C{k-a2} to \C{B}. Since neither \C{Wd1} nor
\C{Wd2} are committed at the beginning, the pre-condition $P$ of the
program is extended with $\neg\committed(\C{Wd1}) \wedge
\neg\committed(\C{Wd2})$, from which $P_1$ follows. The post-condition
$Q_1$ of \C{Wd1} asserts different values for \C{B} depending on
whether or not \C{Wd2} has already committed.  It also asserts that
\C{Wd1} has committed. The post-condition $Q_2$ of \C{Wd2} (not shown)
is similar, but with the roles of \C{Wd1} and \C{Wd2} exchanged. The
post-condition $Q$ follows from $Q_1 \wedge Q_2$, hence valid. It
remains to show that \C{Wd1} satisfies its specification (\C{Wd2} must
also satisfy its spec, but the proof is similar).

First, we focus on the sequential aspect of the proof and show that
the assertions that decorate \C{Wd1} are indeed valid. We consider the
triple associated with the assignment to \C{B} for illustration. The
pre-condition $\phi_4$ asserts different values for \C{B} and \C{v3}
(temporary alias of RHS expression) depending on whether or not
\C{Wd2} is committed. Inside \C{Wd1}, $\committed(\C{Wd2})$ may mean
that either \C{Wd2} happened before \C{Wd1} (hence can be visible to
all of \C{Wd1}), or that it committed concurrently with \C{Wd1} (hence
cannot be visible to all of \C{Wd2}).  Nonetheless, $\phi_4$ only
considers the case when $\C{Wd2} \visar \C{Wd1}$. As we show below,
this is sufficient to prove the post-condition $\phi_5$, which appears
to make a stronger claim about the value of \C{B}. The proof follows
from the RG rule \rulelabel{RG-Asgn} for assignments, which allows us
to easily conclude the following about the execution state after the
assignment:\vspace*{-12pt}

\begin{smathpar}
\begin{array}{l}
    {\neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B} 
    \conj (\neg\committed(\C{Wd2})} \Rightarrow \C{B = k-a1}) \conj\\
    ({\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar
    \C{B}) \conj \\
    ({\committed(\C{Wd2})} \wedge {\C{Wd2} \visar \C{Wd1}} 
    \Rightarrow \C{B = k-a2} \wedge \C{B = k-a2-a1})
\end{array}
\end{smathpar}

\noindent The rule also lets us assume that the execution satisfies trace
invariant ($\I$), which asserts {\sc si} for both transactions.  Since
both transactions now write to \C{B}, {\sc si} requires either
$\C{Wd1} \visar \C{Wd2}$, or $\C{Wd2} \visar \C{Wd1}$. Since \C{Wd1}
is not yet committed ($\neg\committed(\C{Wd1})$), {\sc si} (which
subsumes {\sc rc}) prohibits the former possibility, allowing us the
deduce $\C{Wd2} \visar \C{Wd1}$. This lets us derive
${\committed(\C{Wd2})}  \Rightarrow \C{B = k-a2} \wedge \C{B =
k-a2-a1}$, allowing us to prove the post-condition $\phi_5$.

The second part of the proof is to show that assertions are stable
despite the interference from the the concurrent thread executing
\C{Wd2}. We formalize the interference as a rely relation that
describes the semantics of \C{Wd2} at just the right level of
abstraction. In particular, we would like to capture two important
facts about \C{Wd2}: 
\begin{itemize}
  \item  If \C{Wd2} commits, then it should have already written to
  \C{B}. This fact is needed to show that \C{Wd2} conflicts with
  \C{Wd1}, so that \C{SI(Wd1)} property can be exploited to deduce
  $\C{Wd2} \visar \C{Wd1}$ from $\committed(\C{Wd2})$ in the
  sequential proof.

  \item The value written to \C{B} is either $\C{k-a1-a2}$ or
  $\C{k-a1}$ depending on whether or not \C{Wd1} has already
  committed. This is clearly necessary to deduce the value of \C{B}
  before and after \C{Wd1} commits.
\end{itemize}
 The resultant rely relation ($R_1$) is shown below (note: we
write $(\E'-\E) \Vdash {\committed(\C{Wd2})}$ to express that \C{Wd2}
is not committed in $\E$, but committed in $\E'$.):

\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \neg\underE{\committed(\C{Wd2})} \conj \I(\E) \conj \I(\E')\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{B} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{B=k-a1-a2} \\
%   & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
%       (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
    & & \hspace*{0.8in}\Rightarrow \C{B=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
    & & \hspace*{0.37in} \conj \underE{\committed(\C{Wd1})} \conj
        (\E'-\E) \Vdash {\committed(\C{Wd2})}\\
%       \C{COMMIT(Wd2)} \in (\E'.\A-\E.\A) \\
    & & \hspace*{0.6in}\Rightarrow \C{B=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{B} \\
    & & \hspace*{0.37in}\conj \forall(\eta \in \E'.\A - 
        \E.\A).~\txn(\eta) = \C{Wd2}\}\\
\end{array}
\end{smathpar}
For the sake of completeness, we also explicitly represent the fact
that all the effects added by $R_1$ are of \C{Wd2}. This is needed to
prove that an interference from \C{Wd2} does not affect the local
state of \C{Wd1}.

The stability for $\phi_4$ w.r.t. $R_1$ is trivial to establish: if
$R_1$ does not commit \C{Wd2}, then the value of \C{B} is unaffected,
and if it commits \C{Wd2}, then $\C{Wd2}\not\visar\C{Wd1}$, hence
$\phi_4$ doesn't care. For $\phi_5$, {\sc si} condition should be used
to show that any interference from \C{Wd2} is invalid. The proof
proceeds on the same lines as the Hoare proof discussed above; it is
also discussed in detail in \S\ref{sec:motivation}.
